---
title: "Clear Outcomes Trial"
output: html_document
date: "2023-12-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)

# Set global options to suppress Stan verbose output
options(stan.verbose = FALSE)

# Function to silently load libraries without showing messages in the knitted output
quietly_load_library <- function(library_name) {
  suppressPackageStartupMessages(library(library_name, character.only = TRUE))
}

# Use the function to load required libraries
quietly_load_library("brms")
quietly_load_library("tidybayes")
quietly_load_library("rstan")
quietly_load_library("dplyr")
quietly_load_library("ggplot2")
quietly_load_library("ggridges")
quietly_load_library("ggpubr")
quietly_load_library("ggthemes")
#quietly_load_library("bpp")
quietly_load_library("foreach")
```

### Bayesian Model with Conjugate Multivariate Normal Priors

In the CLEAR outcome trial, the aim was to find a clinically meaninful difference on MACE using Bempedoic Acid in statin-intolerance patients.

Let's first compare the primary outcomes (found significant) using a unimodal Normal shrinkage prior. We take advantage that the logOR is approximately normally distributed, and that the normal distribution has a conjugate (normal) prior. This means that the Posterior and Prior take the same form, and in the case of the normal distribtuion, so does the likelihood!

# Univariate primary outcomes

1.  Skeptical Prior

Definition: A skeptical prior is centered at 0, with the standard deviation calculated such that only 10% of the probability mass is to the left of the Minimal Clinically Important Difference (MCID).

Let MCID be the minimal clinically important difference converted to its log odds ratio equivalent, log_mcid. To find the standard deviation (sigma_skeptical), determine the z-score that leaves 90% of the distribution to the right of log_mcid. This z-score corresponds to the 10th percentile of the standard normal distribution.

```{r}
# Given data from the table
n_treatment <- 6992
events_treatment <- 819
n_control <- 6978
events_control <- 927

# Calculate the odds ratio (OR)
or <- (events_treatment / (n_treatment - events_treatment)) / (events_control / (n_control - events_control))

# Convert OR to logOR
log_or <- log(or)

# Calculate the standard error of the logOR
se_log_or <- sqrt((1 / events_treatment) + (1 / (n_treatment - events_treatment)) + (1 / events_control) + (1 / (n_control - events_control)))


```

Now we need to define the MCID as an absolute risk reduction of 1.3% and convert the absolute risk reduction to logOR for MCID.

```{r}
mcid_arr <- 0.013

# Assuming a baseline risk from the control group
(baseline_risk <- events_control / n_control)

# Calculate the risk in the treatment group
risk_treatment <- baseline_risk - mcid_arr

# Calculate the odds for control and treatment
odds_control <- baseline_risk / (1 - baseline_risk)
odds_treatment <- risk_treatment / (1 - risk_treatment)

# Calculate the odds ratio (OR)
or_mcid <- odds_treatment / odds_control

# For Bayesian analysis, we often use the natural log of the OR
(log_mcid <- log(or_mcid))

```

Here we show that these infact produce desired priors:

```{r}

z_skeptical <- qnorm(0.10)
sigma_skeptical <- abs(log_mcid / z_skeptical)
pnorm(log_mcid,sd=sigma_skeptical)
```

2.  Enthusiastic Prior Definition: An enthusiastic prior is centered at MCID, with the standard deviation calculated such that there is a 30% probability of harm (logOR \> 0).

For the enthusiastic prior, we want 70% of the distribution to be below 0 when the mean is at log_mcid. To find the standard deviation (sigma_enthusiastic), calculate the z-score that corresponds to the 70th percentile of the standard normal distribution.

```{r}
z_enthusiastic <- qnorm(0.70)
sigma_enthusiastic <- abs((0 - log_mcid) / z_enthusiastic)
pnorm(0,mean=log_mcid,sd = sigma_enthusiastic)
```

3.  Pessimistic Prior Definition: A pessimistic prior is centered at -MCID, with the standard deviation calculated such that there is a 30% probability of benefit (logOR \< 0).

Calculation:

For the pessimistic prior, we want 70% of the distribution to be above 0 when the mean is at -log_mcid. To find the standard deviation (sigma_pessimistic), calculate the z-score that corresponds to the 30th percentile of the standard normal distribution.

```{r}
z_pessimistic <- qnorm(0.30)
sigma_pessimistic <- abs((0 + log_mcid) / z_pessimistic)

pnorm(0,mean=-log_mcid,sd = sigma_pessimistic)
```

```{r echo=FALSE}
# Install necessary packages if not already installed
if (!require("metafor")) install.packages("metafor")
library(rstan)
library(metafor)
library(reshape)
library(ggplot2)
library(MASS)
library(ggplot2)
library(reshape2)  # For melting data frames
library(gridExtra)  # 
library(ggpubr)
library(kernlab)

# Load necessary library
library(tidyverse)

# Given data from the table
n_treatment <- 6992
events_treatment <- 819
n_control <- 6978
events_control <- 927

# Calculate the odds ratio (OR)
or <- (events_treatment / (n_treatment - events_treatment)) / (events_control / (n_control - events_control))

# Convert OR to logOR
log_or <- log(or)

# Calculate the standard error of the logOR
se_log_or <- sqrt((1 / events_treatment) + (1 / (n_treatment - events_treatment)) + (1 / events_control) + (1 / (n_control - events_control)))

# Define the MCID as an absolute risk reduction of 1.3%
mcid_arr <- 0.013
# Convert the absolute risk reduction to logOR for MCID
# Assuming a baseline risk from the control group
baseline_risk <- events_control / n_control

# Calculate the risk in the treatment group
risk_treatment <- baseline_risk - mcid_arr

# Calculate the odds for control and treatment
odds_control <- baseline_risk / (1 - baseline_risk)
odds_treatment <- risk_treatment / (1 - risk_treatment)

# Calculate the odds ratio (OR)
or <- odds_treatment / odds_control

# For Bayesian analysis, we often use the natural log of the OR
log_or <- log(or)

# Define the priors based on the specifications

# Skeptical Prior: 90% certainty of no clinically relevant effect, so only 10% of area 
skeptical_sd <- abs(log_mcid / qnorm(0.10))
skeptical_prior <- list(mu = 0, sigma = skeptical_sd)

# Enthusiastic Prior: 30% probability of harm
enthusiastic_sd <- abs(log_mcid) / qnorm(0.70)
enthusiastic_prior <- list(mu = log_mcid, sigma = enthusiastic_sd)

# Pessimistic Prior: 30% probability of benefit
pessimistic_sd <- abs(log_mcid) / qnorm(0.70)
pessimistic_prior <- list(mu = -log_mcid, sigma = pessimistic_sd)

# Output the calculated values and the priors
list(
  log_or = log_or,
  se_log_or = se_log_or,
  skeptical_prior = skeptical_prior,
  enthusiastic_prior = enthusiastic_prior,
  pessimistic_prior = pessimistic_prior
)

```

Conjugate analysis:

```{r}
# Function to calculate the posterior
calc_conjugate_posterior <- function(log_or, se, prior) {
  var_prior <- prior$sigma^2
  var_data <- se^2
  var_post <- 1 / (1 / var_prior + 1 / var_data)
  mu_post <- var_post * (log_or / var_data + prior$mu / var_prior)
  return(list(mu = mu_post, sigma = sqrt(var_post)))
}

# Calculate the posterior for the skeptical prior
posterior_skeptical <- calc_conjugate_posterior(log_or, se_log_or, skeptical_prior)
posterior_skeptical

posterior_enthus <- calc_conjugate_posterior(log_or, se_log_or, enthusiastic_prior)
posterior_enthus

posterior_pess <- calc_conjugate_posterior(log_or, se_log_or, pessimistic_prior)
posterior_pess
```

Show that this approximates the MCMC result:

```{r model-fitting, include=FALSE, message=FALSE, warning=FALSE, error=FALSE}
#
# Install rstanarm if not already installed
library(rstanarm)
# Globally set stan.verbose to FALSE
options(stan.verbose = FALSE)

# Redirect all output to NULL during model fitting
sink(file = "/dev/null", type = "output")

stan_model_code <- "
data {
  real logOR;             // Log odds ratio
  real<lower=0> SE;       // Standard error
  real<lower=0> sd_prior; // Standard deviation of the  prior
  real mu_prior; // mean of the  prior
}
parameters {
  real theta;  // Parameter (log odds ratio) to estimate
}
model {
  // Skeptical prior for theta
  theta ~ normal(mu_prior, sd_prior);
  
  // Likelihood
  logOR ~ normal(theta, SE);
}
"

# Prepare data for Stan, including skeptical_sd
stan_data <- list(logOR = log_or, SE = se_log_or, sd_prior = skeptical_sd,mu_prior=0)

# Run MCMC simulation
fit <- stan(model_code = stan_model_code, data = stan_data, iter = 4000, chains = 4,verbose = FALSE,refresh=0)
```

## Extract the results of the skeptical prior

```{r}
summary(fit)$summary
```

## now for pessimistic

### Prepare data for Stan, including skeptical_sd

```{r}
stan_data <- list(logOR = log_or, SE = se_log_or, mu_prior=pessimistic_prior$mu,sd_prior = pessimistic_prior$sigma)

# Run MCMC simulation
fit <- stan(model_code = stan_model_code, data = stan_data, iter = 4000, chains = 4,verbose = FALSE,refresh=0)

# Extract the results
summary(fit)$summary
```

## For the enthusiastic

### Prepare data for Stan, including skeptical_sd

```{r}
stan_data <- list(logOR = log_or, SE = se_log_or, mu_prior=enthusiastic_prior$mu,sd_prior = enthusiastic_prior$sigma)

# Run MCMC simulation
fit <- stan(model_code = stan_model_code, data = stan_data, iter = 4000, chains = 4,verbose = FALSE,refresh=0)

# Extract the results
summary(fit)$summary
```

Now for the reference prior..

```{r,eval=F,echo=T}
# Load the required packages
library(brms)
library(tidybayes)
library(rstan)
library(dplyr)
library(ggplot2)

# Enter the data from the table into a data frame
meta_data <- data.frame(
  study = c("Goldberg et al., 2019", "Laufs et al., 2019", "Ray et al., 2019", "Nissen et al., 2023"),
  group = rep(c("Bempedoic Acid", "Placebo"), each = 4),
  events = c(32, 9, 68, 831, 21, 0, 42, 952),
  total = c(522, 234, 1487, 6992, 257, 111, 742, 6978)
)

# Prior data from all but the Nissen study
prior_data <- meta_data[meta_data$study != "Nissen et al., 2023",]

# Likelihood data from the Nissen study
likelihood_data <- meta_data[meta_data$study == "Nissen et al., 2023",]

# Formula for meta-analysis with random effects for prior data
prior_formula <- bf(
  events | trials(total) ~ 0 + group + (1|study)
)

# Define priors for the prior data model
# Here you may start with non-informative or weakly informative priors
prior_priors <- c(
  prior(normal(0, 5), class = b, coef = "groupBempedoicAcid"),
  prior(cauchy(0, 0.5), class = sd)
)

# Fit the model to the prior data
prior_model <- brm(
  formula = prior_formula, 
  data = prior_data, 
  family = binomial(),
  prior = prior_priors,
  seed = 123, # for reproducibility
  control = list(adapt_delta = 0.95) # for computational stability
)
# Extract posterior estimates to inform new priors
posterior_estimates <- summary(prior_model)$fixed 

# Extract the mean and standard deviation for the treatment effect (Bempedoic Acid)
bempedoic_effect_mean <- posterior_estimates["groupBempedoic Acid", "Estimate"]
bempedoic_effect_sd <- posterior_estimates["groupBempedoic Acid", "Est.Error"]

# Define the new priors for the full model including Nissen data
new_priors <- c(
  prior(normal(bempedoic_effect_mean, bempedoic_effect_sd), class = b, coef = "groupBempedoic Acid"),
  prior(cauchy(0, 0.5), class = sd) # Assuming heterogeneity prior remains the same
)

full_formula <- bf(
  events | trials(total) ~ 0 + group + (1|study)
)

# Fit the full model including the Nissen study
full_model <- brm(
  formula = full_formula,
  data = meta_data, # Note this is the full dataset including Nissen
  family = binomial(),
  prior = new_priors,
  seed = 123, # for reproducibility
  control = list(adapt_delta = 0.95) # for computational stability
)

# Summarize the full model
summary(full_model)

# Extract the posterior samples for the full model
posterior_full_samples <- posterior_samples(full_model)


```
