---
title: "Clear Outcomes Trial"
output: html_document

date: "2023-12-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)

# Set global options to suppress Stan verbose output
options(stan.verbose = FALSE)

# Function to silently load libraries without showing messages in the knitted output
quietly_load_library <- function(library_name) {
  suppressPackageStartupMessages(library(library_name, character.only = TRUE))
}

# Use the function to load required libraries
quietly_load_library("brms")
quietly_load_library("tidybayes")
quietly_load_library("rstan")
quietly_load_library("dplyr")
quietly_load_library("ggplot2")
quietly_load_library("ggridges")
quietly_load_library("ggpubr")
quietly_load_library("ggthemes")
#quietly_load_library("bpp")
quietly_load_library("foreach")
```

### Bayesian Model with Conjugate Multivariate Normal Priors

In the CLEAR outcome trial, the aim was to find a clinically meaninful difference on MACE using Bempedoic Acid in statin-intolerance patients.

Let's first compare the primary outcomes (found significant) using a unimodal Normal shrinkage prior. We take advantage that the logOR is approximately normally distributed, and that the normal distribution has a conjugate (normal) prior. This means that the Posterior and Prior take the same form, and in the case of the normal distribtuion, so does the likelihood!

# Univariate primary outcomes

1.  Skeptical Prior

Definition: A skeptical prior is centered at 0, with the standard deviation calculated such that only 10% of the probability mass is to the left of the Minimal Clinically Important Difference (MCID).

Let MCID be the minimal clinically important difference converted to its log odds ratio equivalent, log_mcid. To find the standard deviation (sigma_skeptical), determine the z-score that leaves 90% of the distribution to the right of log_mcid. This z-score corresponds to the 10th percentile of the standard normal distribution.

```{r}
# Given data from the table
n_treatment <- 6992
events_treatment <- 819
n_control <- 6978
events_control <- 927

# Calculate the odds ratio (OR)
or <- (events_treatment / (n_treatment - events_treatment)) / (events_control / (n_control - events_control))

# Convert OR to logOR
log_or <- log(or)

# Calculate the standard error of the logOR
se_log_or <- sqrt((1 / events_treatment) + (1 / (n_treatment - events_treatment)) + (1 / events_control) + (1 / (n_control - events_control)))


```

Now we need to define the MCID as an absolute risk reduction of 1.3% and convert the absolute risk reduction to logOR for MCID.

```{r}
mcid_arr <- 0.013

# Assuming a baseline risk from the control group
(baseline_risk <- events_control / n_control)

# Calculate the risk in the treatment group
risk_treatment <- baseline_risk - mcid_arr

# Calculate the odds for control and treatment
odds_control <- baseline_risk / (1 - baseline_risk)
odds_treatment <- risk_treatment / (1 - risk_treatment)

# Calculate the odds ratio (OR)
or_mcid <- odds_treatment / odds_control

# For Bayesian analysis, we often use the natural log of the OR
(log_mcid <- log(or_mcid))

```

Here we show that these infact produce desired priors:

```{r}

z_skeptical <- qnorm(0.10)
sigma_skeptical <- abs(log_mcid / z_skeptical)
pnorm(log_mcid,sd=sigma_skeptical)
```

2.  Enthusiastic Prior Definition: An enthusiastic prior is centered at MCID, with the standard deviation calculated such that there is a 30% probability of harm (logOR \> 0).

For the enthusiastic prior, we want 70% of the distribution to be below 0 when the mean is at log_mcid. To find the standard deviation (sigma_enthusiastic), calculate the z-score that corresponds to the 70th percentile of the standard normal distribution.

```{r}
z_enthusiastic <- qnorm(0.70)
sigma_enthusiastic <- abs((0 - log_mcid) / z_enthusiastic)
pnorm(0,mean=log_mcid,sd = sigma_enthusiastic)
```

3.  Pessimistic Prior Definition: A pessimistic prior is centered at -MCID, with the standard deviation calculated such that there is a 30% probability of benefit (logOR \< 0).

Calculation:

For the pessimistic prior, we want 70% of the distribution to be above 0 when the mean is at -log_mcid. To find the standard deviation (sigma_pessimistic), calculate the z-score that corresponds to the 30th percentile of the standard normal distribution.

```{r}
z_pessimistic <- qnorm(0.30)
sigma_pessimistic <- abs((0 + log_mcid) / z_pessimistic)

pnorm(0,mean=-log_mcid,sd = sigma_pessimistic)
```

```{r echo=FALSE}
# Install necessary packages if not already installed
if (!require("metafor")) install.packages("metafor")
library(rstan)
library(metafor)
library(reshape)
library(ggplot2)
library(MASS)
library(ggplot2)
library(reshape2)  # For melting data frames
library(gridExtra)  # 
library(ggpubr)
library(kernlab)

# Load necessary library
library(tidyverse)

# Given data from the table
n_treatment <- 6992
events_treatment <- 819
n_control <- 6978
events_control <- 927

# Calculate the odds ratio (OR)
or <- (events_treatment / (n_treatment - events_treatment)) / (events_control / (n_control - events_control))

# Convert OR to logOR
log_or <- log(or)

# Calculate the standard error of the logOR
se_log_or <- sqrt((1 / events_treatment) + (1 / (n_treatment - events_treatment)) + (1 / events_control) + (1 / (n_control - events_control)))

# Define the MCID as an absolute risk reduction of 1.3%
mcid_arr <- 0.013
# Convert the absolute risk reduction to logOR for MCID
# Assuming a baseline risk from the control group
baseline_risk <- events_control / n_control

# Calculate the risk in the treatment group
risk_treatment <- baseline_risk - mcid_arr

# Calculate the odds for control and treatment
odds_control <- baseline_risk / (1 - baseline_risk)
odds_treatment <- risk_treatment / (1 - risk_treatment)

# Calculate the odds ratio (OR)
or <- odds_treatment / odds_control

# For Bayesian analysis, we often use the natural log of the OR
log_or <- log(or)

# Define the priors based on the specifications

# Skeptical Prior: 90% certainty of no clinically relevant effect, so only 10% of area 
skeptical_sd <- abs(log_mcid / qnorm(0.10))
skeptical_prior <- list(mu = 0, sigma = skeptical_sd)

# Enthusiastic Prior: 30% probability of harm
enthusiastic_sd <- abs(log_mcid) / qnorm(0.70)
enthusiastic_prior <- list(mu = log_mcid, sigma = enthusiastic_sd)

# Pessimistic Prior: 30% probability of benefit
pessimistic_sd <- abs(log_mcid) / qnorm(0.70)
pessimistic_prior <- list(mu = -log_mcid, sigma = pessimistic_sd)

# Output the calculated values and the priors
list(
  log_or = log_or,
  se_log_or = se_log_or,
  skeptical_prior = skeptical_prior,
  enthusiastic_prior = enthusiastic_prior,
  pessimistic_prior = pessimistic_prior
)

```

Conjugate analysis:

```{r}
# Function to calculate the posterior
calc_conjugate_posterior <- function(log_or, se, prior) {
  var_prior <- prior$sigma^2
  var_data <- se^2
  var_post <- 1 / (1 / var_prior + 1 / var_data)
  mu_post <- var_post * (log_or / var_data + prior$mu / var_prior)
  return(list(mu = mu_post, sigma = sqrt(var_post)))
}

# Calculate the posterior for the skeptical prior
posterior_skeptical <- calc_conjugate_posterior(log_or, se_log_or, skeptical_prior)
posterior_skeptical

posterior_enthus <- calc_conjugate_posterior(log_or, se_log_or, enthusiastic_prior)
posterior_enthus

posterior_pess <- calc_conjugate_posterior(log_or, se_log_or, pessimistic_prior)
posterior_pess
```

Show that this approximates the MCMC result:

```{r model-fitting, include=FALSE,message=FALSE, warning=FALSE, error=FALSE,echo=TRUE}
#
# Install rstanarm if not already installed
library(rstanarm)
# Globally set stan.verbose to FALSE
options(stan.verbose = FALSE)

# Redirect all output to NULL during model fitting
sink(file = "/dev/null", type = "output")

stan_model_code <- "
data {
  real logOR;             // Log odds ratio
  real<lower=0> SE;       // Standard error
  real<lower=0> sd_prior; // Standard deviation of the  prior
  real mu_prior; // mean of the  prior
}
parameters {
  real theta;  // Parameter (log odds ratio) to estimate
}
model {
  // Skeptical prior for theta
  theta ~ normal(mu_prior, sd_prior);
  
  // Likelihood
  logOR ~ normal(theta, SE);
}
"

# Prepare data for Stan, including skeptical_sd
stan_data <- list(logOR = log_or, SE = se_log_or, sd_prior = skeptical_sd,mu_prior=0)

# Run MCMC simulation
fit <- stan(model_code = stan_model_code, data = stan_data, iter = 4000, chains = 4,verbose = FALSE,refresh=0)
```

## Extract the results of the skeptical prior

```{r}
summary(fit)$summary
```

## now for pessimistic

### Prepare data for Stan, including skeptical_sd

```{r}
stan_data <- list(logOR = log_or, SE = se_log_or, mu_prior=pessimistic_prior$mu,sd_prior = pessimistic_prior$sigma)

# Run MCMC simulation
fit <- stan(model_code = stan_model_code, data = stan_data, iter = 4000, chains = 4,verbose = FALSE,refresh=0)

# Extract the results
summary(fit)$summary
```

## For the enthusiastic

### Prepare data for Stan, including skeptical_sd

```{r}
stan_data <- list(logOR = log_or, SE = se_log_or, mu_prior=enthusiastic_prior$mu,sd_prior = enthusiastic_prior$sigma)

# Run MCMC simulation
fit <- stan(model_code = stan_model_code, data = stan_data, iter = 4000, chains = 4,verbose = FALSE,refresh=0)

# Extract the results
summary(fit)$summary
```

## Now for the reference prior, we will perform a Bayesian meta analysis, and then use the results in both a conjugate and MCMC analysis to show the results match.


\item A separate baseline risk of CAD for each study (via factor(study)).
\item A common treatment effect of Bempedoic acid across all studies (via the bpd term).
\item Variation in the treatment effect of Bpd across different studies (via the random effect (bpd - 1|study)).

First, we will conduct a meta-analysis of the 3 RCTs that tested similar filter-based devices. This meta-analysis will form the basis of our subsequent prior.

```{r,eval=T,echo=T}
#Load the packages & functions that we will need for this analysis
require("brms")
require("tidybayes")
detach("package:rstanarm", unload = TRUE)
detach("package:rstan", unload = TRUE)
require("dplyr")
require("ggplot2")
require("ggridges")
require("ggpubr")
require("ggthemes")
require("bpp")
require("foreach")



#Trial 1: goldberg et al
goldberg <- data.frame(
  study = "goldberg",
  #Trial name
  total_n = c(522, 257),
  #Total sample size in each group
  events_n = c(32, 21),
  #Patients with MACE in each group
  bpd = c(0.5,-0.5) #bpd (Yes/No)
)

#Trial 2: laufs et al
laufs <- data.frame(
  study = "laufs",
  #Trial name
  total_n = c(234, 111),
  #Total sample size in each group
  events_n = c(9, 0),
  #Patients with MACE in each group
  bpd = c(0.5,-0.5) #bpd 1/0 (Yes/No)
)


#Trial 3: Ray
ray <- data.frame(
  study = "ray",
  #Trial name
  total_n = c(1487, 742),
  #Total sample size in each group
  events_n = c(68, 42),
  #Patients with MACE in each group
  bpd = c(0.5,-0.5) #bpd 1/0 (Yes/No)
)


#Combine the data from the 6 trials into a single dataframe
combined_data <- rbind(goldberg, ray, laufs)

#Now, we will write the formula we will use in the regression model
ma_formula <- bf(events_n |
                   trials(total_n) ~ 0 + #Remove the intercept from the model (such that the risk of stroke is not modeled using a common term but is modeled separately for each study)
                   factor(study) + #A term to refer to each study
                   bpd + # Fixed treatment effect of bpd
                   (bpd - 1 |study)) # We allow for a random slope, but a fixed intercept effect
                 
```


Now, we will set up our priors (on the log-odds scale), as we recall that the natural link for a binomial process is the logit function, and the logOR is normally distributed.

\item Treatment effect prior, is `flat' suggesting that all the information will be gelamed from the observed studies which will inform our later posterior

\item Fixed effect prior for MACE within the population, which is centered on the logOR of 13% (-2) that we observed previously

\item Heterogeneity prior, a 'fat tailed' cauchy prior with sd 0.5, commonly chosen in meta analyses

```{r,eval=T,echo=TRUE}
#Prior for the Treatment effect of BPD (a normal distribution with a mean of 0 and a standard deviation of 10 implies a flat prior (we are unsure if the prior effect is small or large)
bpd_prior <- prior(normal(0, 10), class = b, coef = "bpd")

#Prior for the Baseline risk of MACE (this prior is centered on a baseline risk of CAD of 13.%, recall from th baseline analysis: log(baseline_risk))
## really thing here if each study has it's own fixed effect, we need to specify this (or allow for a random MACE effect )

#The stroke_prior that you mentioned, prior(normal(-3.2, 0.5), class = b), is indeed for the fixed effects. But without specifying coef, it is not clear whether this prior is meant for the intercepts created by factor(study) or for another fixed effect. If you want this prior to be applied to the intercepts for each study (which you've modeled as fixed effects by using factor(study)), you would need to clarify that in the prior specification, typically by specifying each coefficient if they are to be treated differently.
# #stroke_prior <- c(set_prior("normal(-3.2, 0.5)", class = "b", coef = "study1"),
#                   set_prior("normal(-3.2, 0.5)", class = "b", coef = "study2"),
#                   ... and so on for each study ...)

mace_prior <- prior(normal(-2, 0.5), class = b)

#Prior for between-trial Heterogeneity (this prior expresses our uncertainty about how similar trials are to one another in terms of the effectiveness of bpd. A half-cauchy with a scale parameter of 0.5 represents a common choice in most meta-analyses)
heterogeneity_prior <- prior(cauchy(0, 0.5), class = sd)

#Now, we will collect these priors together and store them in "ma_priors" (ma: meta-analysis)
ma_priors <- c(bpd_prior, mace_prior, heterogeneity_prior)

#Because Bayesian analyses can be somewhat time-consuming, we will use all the cores our machine has to increase computational speed
options(mc.cores = parallel::detectCores())


ma_model <-
  brm(
    data = combined_data,
    #Use the combined dataset for the 6 trials
    family = binomial,
    #Using a binomial distribution (Since our outcome is a binary MACE/noMACE
    formula = ma_formula,
    #Use our formula
    seed = 100,
    #Set seed for reproducibility
    prior = ma_priors,
    #Use our prior
    control = list(adapt_delta = 0.99) 
  )
saveRDS(ma_model,"~/Library/CloudStorage/Dropbox-Personal/ma_model.rds")
```

```{r}
ma_model=readRDS("~/Library/CloudStorage/Dropbox-Personal/ma_model.rds")

sum_ma_model <- summary(ma_model)

#Get the parameters of the row which corresponds to the treatment (BPD)
ma_bpd_par <- sum_ma_model$fixed[rownames(sum_ma_model$fixed) == "bpd", ]

##Store relevant variables
#The log odds ratio
ma_bpd_lnor <- round(ma_bpd_par$Estimate, 2)
#Standard error of the log odds ratio
ma_bpd_lnor_sd <- round(ma_bpd_par$Est.Error, 2)
#The odds ratio
ma_bpd_or <- round(exp(ma_bpd_par$Estimate), 2)
#The lower limit of the 95% credible interval of the odds ratio
ma_bpd_or_lci <- round(exp(ma_bpd_par$`l-95% CI`), 2)
#The upper limit of the 95% credible interval of the odds ratio
ma_bpd_or_uci <- round(exp(ma_bpd_par$`u-95% CI`), 2)

#Sample the posterior distribution of study-level estimates and the overall estimate of treatment effect
study_es <- ma_model %>%
  spread_draws(r_study[study, ], b_bpd) %>%
  mutate(b_bpd = r_study + b_bpd, #Create treatment effect estimates for each study
         type = "Study-level estimate") #Clarify that this is the treatment effect for each study


pooled_es <- spread_draws(ma_model, b_bpd) %>% 
  mutate(study = "Overall Effect Size", #Clarify that this is the pooled/overall treatment effect
         type = "Pooled estimate") #Same

#Exponentiate to get odds instead of log-odds ratio
fp_data <- bind_rows(study_es, pooled_es) %>%
  mutate(b_bpd = b_bpd %>% exp)
```


```{r plotting,echo=F}

#Create title and subtitles for the plot
main_title <- "Figure 1: Creating a meta-analysis-based prior for the effecitveness of BPD in preventing MACE"
subtitle <- "This analysis synthesizes the 3 RCTs constituting the RCT evidence base for BPD prior to the CLEAR Trial."


#Plot
ggplot(data = fp_data,
       aes(y = study,
           x = b_bpd,
           fill = type
       )) +
  #Add Density plots
  geom_density_ridges(col = NA,
                      scale = 0.9, #Slightly decrease size so it fits
                      alpha = 0.7,
  ) +
  geom_vline(xintercept = 1, color = "black", 
             lwd = 1, linetype = 2) +
  #Set colors
  scale_fill_manual(values = c("salmon", "lightblue")) +
  #Create title
  ggtitle(main_title,
          subtitle = subtitle) +
  #X and Y axes aesthetics
  scale_y_discrete(name = "Study") +
  scale_x_continuous(name = "Odds Ratio",
                     trans = "log",
                     breaks = c(0.25, 0.5, 1, 2, 4)) +
  #Set reasonable Y axis limits
  coord_cartesian(xlim = c(0.25, 4)) +
  #Set theme
  theme_pubclean() +
  theme(text = element_text(size = 23),
        plot.title=element_text(face = "bold",hjust = 0.0, size = 20),
        plot.subtitle = element_text(face = "bold", size = 15, hjust = 0.0, color = "grey45"),
        axis.text.x = element_text(size = 20, face = "bold"),
        axis.text.y = element_text(size = 15, face = "bold"),
        axis.title.x = element_text(size = 25, face = "bold"),
        axis.title.y = element_blank(),
        axis.line = element_line(colour = "black", linewidth = 1.2),
        plot.margin = margin(0.5, 1, 0.5, 1, "cm"),
        legend.background = element_rect(fill = "transparent"),
        legend.position = "none",
        legend.text = element_text(size = 12, face = "bold"),
        legend.key.width = unit(1.5, "cm"),
        legend.key.height = unit(0.75, "cm"))

```

## Now let's use this as a prior

We will show with MCMC and with conjugate normal

```{r message=FALSE, warning=FALSE, error=FALSE,echo=TRUE,eval=TRUE}

# Enter the data from the table into a data frame
meta_data <- data.frame(
  study = c("Goldberg et al., 2019", "Laufs et al., 2019", "Ray et al., 2019", "Nissen et al., 2023"),
  group = rep(c("Bempedoic Acid", "Placebo"), each = 4),
  events = c(32, 9, 68, 831, 21, 0, 42, 927),
  total = c(522, 234, 1487, 6992, 257, 111, 742, 6978)
)


# Likelihood data from the Nissen study
likelihood_data <- meta_data[meta_data$study == "Nissen et al., 2023",]

ma_bpd_lnor <- round(ma_bpd_par$Estimate, 2)
#Standard error of the log odds ratio
ma_bpd_lnor_sd <- round(ma_bpd_par$Est.Error, 2)

# Convert 'group' to a factor and create a binary treatment indicator
likelihood_data$bpd <- ifelse(likelihood_data$group == "Bempedoic Acid", 0.5, -0.5)

library(brms)
# Define the prior
# Define the mean and standard deviation for the log OR
log_or_mean <- ma_bpd_lnor  # Mean log OR
log_or_sd <- ma_bpd_lnor_sd    # Standard deviation of the log OR

# Assuming 'bpd' is the name of the variable for the treatment effect in your model
# Set the prior using the mean and standard deviation
nissen_prior <- set_prior(
  paste("normal(", log_or_mean, ",", log_or_sd, ")"), 
  class = "b", 
  coef = "bpd"
)

# Verify the prior
nissen_model <- brm(
  formula = bf(events | trials(total) ~ bpd),  # Update 'bpd' if your variable name is different
  data = likelihood_data,
  family = binomial(),
  prior = nissen_prior,
  seed = 102,
  control = list(adapt_delta = 0.95)
)

saveRDS(nissen_model,"~/Library/CloudStorage/Dropbox-Personal/nissen_model.rds")
```


# Summary of the Nissen study model

```{r}
nissen_model=readRDS("~/Library/CloudStorage/Dropbox-Personal/nissen_model.rds")
summary(nissen_model)
```

Now let's show that this matches with our conjugate analysis

```{r}

# Given data from the table
n_treatment <- 6992
events_treatment <- 819
n_control <- 6978
events_control <- 927

# Calculate the odds ratio (OR)
or <- (events_treatment / (n_treatment - events_treatment)) / (events_control / (n_control - events_control))

# Convert OR to logOR
log_or <- log(or)

# Calculate the standard error of the logOR
se_log_or <- sqrt((1 / events_treatment) + (1 / (n_treatment - events_treatment)) + (1 / events_control) + (1 / (n_control - events_control)))


calc_conjugate_posterior(log_or = log_or,se = se_log_or,prior = list(mu=ma_bpd_lnor,sigma=ma_bpd_lnor_sd))

```


